|

### 2019-11-30 analysis/paper/cover_letter.Rmd
Prep material for submission to CJFAS

Howard I. Browman, Ph.D.  
Editor-in-Chief  
ICES Journal of Marine Science  
Institute of Marine Research  
Marine Ecosystem Acoustics Group  
Austevoll Research Station  
5392 Storebø  
Norway

@@

Dr. Browman,

I am writing to submit our manuscript entitled, **"Improving the communication and accessibility of stock assessment using interactive visualization tools"** for consideration as a **_Quo Vadimus_** article in **ICES Journal of Marine Science**. In our manuscript, we discuss the merits of applying modern interactive visualization tools to stock assessment processes and we hope to convince readers that these tools, which we use every day on a range of web sites, are very accessible to practicing biologists. We argue that the application of these tools alleviates some of the difficulty in presenting the mass of information produced by complex monitoring programs and statistical models to end-users in a meaningful, interactive and engaging way. To exemplify this point, we describe the structure of three interactive dashboards developed to aid the exploration and communication of results from 1) a fishery-independent trawl survey, 2) a mark-recapture tagging program, and 3) a stock assessment model. In each case we discuss the efficacy of applying such tools in a stock assessment context and contrast this approach with more traditional modes of communicating the results from complex monitoring programs and stock assessment models.

We believe that this is a fitting and timely article for the Journal's **_Quo Vadimus_** type as there are a growing number of scientists from a wide range of disciplines applying these tools to explore and communicate their results. We can personally attest that the application of interactive tools has helped us in our daily work and we hope that our perspective will appeal to readers of **ICES Journal of Marine Science** and encourage the application of interactive tools.

>>>>

Editorial office  
Canadian Journal of Fisheries and Aquatic Sciences  
Canadian Science Publishing (NRC Research Press)  
65 Auriga Drive, Suite 203   
Ottawa, ON K2E 7W6  
Canada

@@

Dear Editors,

I am writing to submit our manuscript entitled, **"Improving the communication and accessibility of stock assessment using interactive visualization tools"** for consideration as a **_Perspectives_** article in **Canadian Journal of Fisheries and Aquatic Sciences**. In our manuscript, we discuss the merits of applying modern interactive visualization tools to stock assessment processes and we hope to convince readers that these tools, which we use every day on a range of web sites, are very accessible to practicing biologists. We argue that the application of these tools alleviates some of the difficulty in presenting the mass of information produced by complex monitoring programs and statistical models to end-users in a meaningful, interactive and engaging way. To exemplify this point, we describe the structure of three interactive dashboards developed to aid the exploration and communication of results from 1) a fishery-independent trawl survey, 2) a mark-recapture tagging program, and 3) a stock assessment model. In each case we discuss the efficacy of applying such tools in a stock assessment context and contrast this approach with more traditional modes of communicating the results from complex monitoring programs and stock assessment models.

We believe that this is a fitting and timely article for the Journal's **_Perspectives_** type as there are a growing number of scientists from a wide range of disciplines applying these tools to explore and communicate their results. We can personally attest that the application of interactive tools has helped us in our daily work and we hope that our perspective will appeal to readers of **Canadian Journal of Fisheries and Aquatic Sciences** and encourage the application of interactive tools.

### 2019-11-30 analysis/paper/manuscript.Rmd
Prep material for submission to CJFAS

---
title: |
  Improving the communication and accessibility of stock assessment using interactive visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Robert Rogers, Keith P. Lewis"
date: |
  Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
csl: ices.csl
abstract: |
  **Running title**: Interactive stock assessment

@@

  **Abstract**: Scientists across many fields are faced with the challenge of synthesizing and communicating information from large and complex data sets. The field of stock assessment is no exception as the volume and variety of the data has grown alongside the computational methods used to integrate them. While this growth in data and model complexity has improved the assessments of many stocks, the process of communicating the results to colleagues, stakeholders and fisheries managers in a meaningful way has become more daunting. The traditional approach of presenting information across a series of static slides often fails to convey the richness of information available and, as such, important patterns and details are easily overlooked. Here we contend that this problem can be mediated through the effective use of new open source tools for building interactive visualizations. These tools allow a broader audience to conduct detailed explorations of the results, leading to a deeper and collective understanding of both the data and models used to inform stock assessments. As a consequence, the peer review process is more open and accessible and the resulting science advice is improved and widely supported.

@@

The data sets used in stock assessments are constantly growing due to the continuation of long-term monitoring efforts, the addition of new monitoring programs, or both. As such, stock assessment biologists need to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from "fishery-independent" surveys that track changes in abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are complex and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. It has long been recognized that this challenge can be mediated, to a degree, by the application of interactive visualization tools as they allow detailed exploration of the data behind a plot [@tukey1985; @fisherkeller1988]. For instance, the ability to zoom in on features or areas of interest, turn off layers and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of data. These abilities are illustrated  by two recently developed interactive tools: 1) a tool designed to quickly examine fishery-independent survey data; and 2) an interactive  mapping tool developed for the exploration of a long-term tagging study.

@@

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model reduces thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The traditional approach would involve producing static presentations and documents with a series of figures and tables.  However, with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved for at least two reasons. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

@@

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Catch"), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" page, trends from different models are compared under the "Comps" page, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents, however, there are two key benefits of this approach over the *modus operandi* of producing static documents and slides: 1) interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips); and, 2) the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables and values into documents and slides. Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and comparisons of models with different configurations. The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of both data and the model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussion helped the assessment biologists, managers and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our three examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but all had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Finally, scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas.

@@

We thank the numerous colleagues and participants of various stakeholder and stock assessment meetings who encouraged us to further develop these interactive visualization tools, and especially those who took the time to make suggestions on how to make them more accessible and useful. We are also grateful for the constructive feedback from Gary Carvalho, Kelli Johnson, Colin Millar and two anonymous reviewers.

>>>>

---
title: |
  Improving the communication and accessibility of stock assessment using interactive visualization tools
author: "Paul M. Regular, Gregory J. Robertson, Robert Rogers, Keith P. Lewis"
date: |
  Fisheries and Oceans Canada, Northwest Atlantic Fisheries Center, 80 East White Hills, St. John's, Newfoundland and Labrador, A1C 5X1, Canada
output: 
  bookdown::word_document2:
    reference_docx: template.docx
bibliography: references.bib
csl: cjfas.csl
abstract: |
  **Running title**: Interactive stock assessment

@@

  **Abstract**: Scientists across many fields are faced with the challenge of synthesizing and communicating information from large and complex data sets. The field of stock assessment is no exception as the volume and variety of the data has grown alongside the computational methods used to integrate them. While this growth in data and model complexity has improved the assessments of many stocks, the process of communicating results from multiple monitoring programs and integrated assessment models to colleagues, stakeholders and fisheries managers in a meaningful way has become more daunting. The traditional approach of presenting information across a series of static slides often fails to convey the richness of information available and, as such, important patterns and details are easily overlooked. Here we contend that this problem can be mediated through the effective use of new open source tools for building interactive visualizations. These tools allow a broader audience to conduct detailed explorations of the results, leading to a deeper and collective understanding of both the data and models used to inform stock assessments. As a consequence, the results of the peer review process are more open and accessible.

@@

The data sets used in stock assessments are constantly growing due to the continuation of long-term monitoring efforts, the addition of new monitoring programs, or both. As such, stock assessment biologists need to manage large volumes of data from a variety of sources. Time series of reported landings and catch-at-age are often analyzed in conjunction with data from "fishery-independent" surveys that track changes in relative abundance and, in many cases, also monitor trends in biological factors such as age composition, growth rates, sex ratios and maturation stages. For some data-rich stocks, mark and recapture studies are also carried out to estimate movement, migration, growth rate, natural mortality, and discard mortality. All of the above-mentioned data sets are complex and as the volume and variety of these data increases, it becomes more difficult to be aware of the details of each data set and discover key patterns within each. It has long been recognized that this challenge can be mediated, to a degree, by the application of interactive visualization tools as they allow detailed exploration of the data behind a plot [@tukey1985; @fisherkeller1988]. For instance, the ability to zoom in on features or areas of interest, turn off layers and hover over specific points to reveal more information creates an interactive user-driven experience that expedites explorations of data. These abilities are illustrated  by two recently developed interactive tools: 1) a tool designed to quickly examine fishery-independent survey data; and 2) an interactive  mapping tool developed for the exploration of a long-term tagging study.

@@

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model simplifies thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The traditional approach would involve producing static presentations and documents with a series of figures and tables.  However, with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved for at least two reasons. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

@@

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Catch"), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" page, trends from different models are compared under the "Comps" page, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents. However, there are two key benefits of this approach over the modus operandi of producing static documents and slides.  First, interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips).  Second, the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables and values into documents and slides, although, we note that this point is not exclusive to the proposed approach (e.g. slide shows can easily be made using R markdown). Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and comparisons of models with different configurations. The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of both data and the model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussions left us with the impression that this tool helped the assessment biologists, academics, resource managers and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables. Feedback and personal testimony from multiple stakeholders who have used the dashboard reaffirmed our impression that the tool is helpful as there was general agreement that access to the details eases communication and relieve concerns that something is hidden.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our three examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but all had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. There are also a growing number of exemplars (Appendix A), beyond what we show here, to inspire the development of new interactive tools. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding.

@@

We thank the numerous colleagues and participants of various stakeholder and stock assessment meetings who encouraged us to further develop these interactive visualization tools, and especially those who took the time to make suggestions on how to make them more accessible and useful. We are also grateful for the constructive feedback from Gary Carvalho, Kelli Johnson, Colin Millar and four anonymous reviewers.
